---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# phylotax

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/brendanf/phylotax.svg?branch=master)](https://travis-ci.com/brendanf/phylotax)
[![Codecov test coverage](https://codecov.io/gh/brendanf/phylotax/branch/master/graph/badge.svg)](https://codecov.io/gh/brendanf/phylotax?branch=master)
<!-- badges: end -->

## Installation

Install the development version from [GitHub](https://github.com/) with:

```r
# install.packages("devtools")
devtools::install_github("brendanf/phylotax")
```

## Usage

The PHYLOTAX algorithm takes as input taxonomic annotations from one or more
primary taxonomic assignment algorithms, and refines them using a taxonomic tree.
The refinements are of two basic types:

  1. Tips of the tree which are unassigned are assigned to a taxon if the tree
     supports their inclusion in that taxon.
  2. Conflicting assignments between multiple primary algorithms are resolved
     using the tree.
     
The `phylotax` package also includes the wrapper function `taxonomy()` which
assigns taxonomy to sequences using
[DADA2](https://benjjneb.github.io/dada2/assign.html),
[IDTAXA](http://www2.decipher.codes/Classification.html),
or [SINTAX](https://www.drive5.com/usearch/manual/cmd_sintax.html), along the
`taxtable()` function which converts the results to a uniform format.

### Example data

Here is an example of a tree:

```{r, fig.asp = 0.5}
library(phylotax)
plot(example_tree(), show.node.label = TRUE)
```

Here is a set of taxonomic assignments for the tips of the tree, based on
two hypothetical primary assignment algorithms "XTAX" and "YTAX".
The required columns are "label", "rank", and "taxon", which identify individual OTUs,
taxonomic ranks, and taxonomic assignments.
The OTU labels are the same as the tip labels on the tree,
in this case the letters A-F, although some are missing from the taxonomy table
because the algorithms could not place them.
Only assignments at the rank of genus are present in this small example,
and the genera in question are called "Tax1" and "Tax2".
Our example also has a "method" column, which PHYLOTAX uses to identify which
assignments come from the same source.

```{r}
example_taxa()
```

If we sort by the taxon label, we can see that XTAX and YTAX disagree about the
assignment of tips B and D.
Neither algorithm has placed tips A and E, and XTAX also failed to place tip F. 

```{r}
dplyr::arrange(example_taxa(), label)
```

### Use PHYLOTAX

Use PHYLOTAX to resolve conflicts and assign additional tips where the tree
supports it.

```{r}
phylotax_out <- phylotax(tree = example_tree(), taxa = example_taxa())
```

PHYLOTAX returns a list of class "`phylotax`" containing the tree,
taxa assignments for tips and internal nodes, as well as tables dividing the
primary assignments into those which were rejected, those which were retained,
and those which were missing from the input tree.

```{r}
phylotax_out$assigned
```

```{r}
phylotax_out$retained
```

```{r}
phylotax_out$rejected
```


PHYLOTAX has used the following logic:

1. It's not possible to decide what the root (node 1) is, because one of its
  direct children (tip A) is completely unassigned.
2. It's not possible to decide node 2, because there are differences between
  assignments for two of its descendents (tip C and tip F).
3. All of the descendents of node 3 (tip B and tip C) have at least one
  assignment of Tax2. PHYLOTAX removes all conflicting assignments (XTAX's
  assignment of Tax1 to tip B) and gives its own assignment of Tax2 to node 3
  and all its children.
4. All of the descendents of node 4 (tips D, E, and F) either have an assignment
  of Tax1 (D and F) or are unassigned (E).  Furthermore, both branches coming
  from node 4 (tip D and node 5) do have some assignments.  PHYLOTAX removes
  the conflicting assignments (XTAX's assignment of Tax2 to tip D) and gives its
  own assignment of Tax1 to node 4 and all its children.
5. At node 5, there is nothing to do, because PHYLOTAX already assigned it to
  Tax1 in step 4.

### Continued analysis

If you are continuing on with analysis using the
[phyloseq](https://joey711.github.io/phyloseq/index.html) package, then you can
easily create a `phyloseq` object from a `phylotax` object and an OTU table.

To demonstrate, we first create a random presence/absence OTU table for 5
samples and the 6 OTUS present in our example dataset.
Each species has a 50% chance to be present in each sample.

```{r}
library(phyloseq)
set.seed(1)
otus <- matrix(rbinom(30, 1, 0.5),nrow = 5, dimnames = list(1:5, LETTERS[1:6]))
otus <- otu_table(otus, taxa_are_rows = FALSE)
```

Now we can easily create the phyloseq object.

```{r}
physeq <- phylotax_to_phyloseq(phylotax_out, otus)
```

And use it for plots or whatever further analysis is needed.

```{r, fig.asp = 0.25}
plot_tree(physeq, color = "genus")
```

